Using SMT engine to generate Symbolic Automata\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2018-01-09 2017-Xudong-SMT/FM-2018-extended.tex
Version initiale pour soumission FM-extended

\begin{abstract}
  %% \TODO{not good for a conference, (to be rewriten by Eric)}
  %% We implement the symbolic semantics of open pNets using their
  %% so-called ``Open Automaton'' behavioral semantics. This involve
  %% building predicates expressing the synchronization conditions
  %% allowing some combination of events in the pNet system. These
  %% predicates are typically built using first order logic, plus some
  %% predicates specific of particular action algebras. To reduce the
  %% complexity of the generated open automata, we use the Z3 SMT engine
  %% to check satisfiability of the predicates, and prune the state
  %% space..

@@

\paragraph{Contribution}
In the article we show how:
\begin{itemize}
\item We define the open automaton generation algorithm, and we
  implemented a full working prototype, within the 
  VerCors platform. In the process, we improved the
  semantics rules from~\cite{henrio:Forte2016}, and add features in
  the algorithm to deal the full 
  model, including management of variables and assignments.
%  \item We extended the meta-model of action expressions, and the
%    associated API, in the VerCors platform, to be able to manage
%    parameterisation over the action algebra operators.
\item We implement the interaction between our algorithm and the Z3
    SMT solver, for checking satisfiability of the transitions
    generated by the algorithm.
\item We validate the whole approach on small use-cases covering all
      important features of the model.
\end{itemize}

@@

% \TODO{I have removed some of the technical stuff, adding referemces to Forte'16.
%  We will see if this is enough, and will have to check if this is coherent, and if we can remove more, but for ICTAC, we MUST keep enough formal stuff}

@@

%% \TODO{[We haven't used the input variables in this paper, we can remove them.]: NO, first see what we use to explain CCS use-case, then leave only a short definition of what we need}
%% \QIN{[To be able to reason about the data flow between pLTSs, we
%% distinguish \emph{input variables} of the form $?x$ within terms; the function
%% $\vars(t)$ identifies the set of variables in a term
%% $t\in\AlgT$, and $iv(t)$ returns its input variables.]}

pNets can encode naturally the notion of input actions as found e.g. in value-passing CCS
\cite{Milner89} or of usual point-to-point message passing calculi, but it also allows
for more general mechanisms, like gate negotiation in Lotos, or broadcast
communications.
%\TODO{I removed the CCS input action stuff, check later if we need to
%  replace it here by something different about input action notation}.

%% Formally, value-passing actions \emph{\`a la} CCS, denoted as
%% ``?ch(x)'' and ``!ch(exp)'' in our CCS running example, stand
%% respectively for $CCSact(i(ch),?x)$ and $CCSact(ch,exp)$, where $ch$
%% is a variable denoting a channel name, $?x$ is an input variable
%% denoting the data argument received as input on $ch$, and $exp$ some
%% data expression (containing no input variables). In our example we
%% will use the usual CCS syntax. 
%% %For Lotos-style
%% %distributed synchronization, we use synchronization vectors that encode matching data
%% %offers on a common gate; or 
%% We can also use more complex action structure such as Meije-SCCS action
%% monoids, like in $a.b$, $a^{f(n)}$ (see \cite{deSimone85}). The expressiveness of the synchronization constructs
%% will depend on the action algebra.

@@

\begin{definition}[pLTS]
\label{pLTS}
A pLTS is a tuple
$pLTS\triangleq\mylangle S,s_0, \to\myrangle$ where:
\begin{itemize}
\item[$\bullet$]
$S$ is a set of states.
\item[$\bullet$]
$s_0 \in S$ is the initial state.
%\item[$\bullet$]
 %Variables in
%$\iv(\alpha)$ are assigned by the action, other variables can be assigned
%by the additional assignments.
\item[$\bullet$] $\to \subseteq S \times L \times S$ is the transition relation and 
$L$ is the set of labels of the form
$\langle \alpha,~e_b,~(x_j\!:= {e}_j)^{j\in J}\rangle$,
where $\alpha \in\AlgA$ is a parameterized action, $e_b \in
\AlgB$ is a guard, and the variables $x_j\in P$
are assigned the expressions $e_j\in \AlgE$.
If 
%% $s \xrightarrow{\langle \alpha,~e_b,~(x_j\!:= {e}_j)^{j\in
%% 		J}\rangle} s'\in \to $ then $\iv(\alpha)\!\subseteq\! \vars(s')$, 
%% 		$\vars(\alpha)\backslash \iv(\alpha)\!\subseteq\! \vars(s)$, 
%% 		$\vars(e_b)\!\subseteq\! \vars(s')$, and
%% 		$\forall j\!\in\! J .\,\vars(e_j)\!\subseteq\! \vars(s)\land 
%% x_j\!\in\!\vars(s')$.

$s \xrightarrow{\langle \alpha,~e_b,~(x_j\!:= {e}_j)^{j\in
		J}\rangle} s'\in \to $ then 
		$\vars(\alpha)\!\subseteq\! \vars(s)$, 
		$\vars(e_b)\!\subseteq\! \vars(s')$, and
		$\forall j\!\in\! J .\,\vars(e_j)\!\subseteq\! \vars(s)\land 
x_j\!\in\!\vars(s')$.

%,  and $s= s'\Rightarrow J=\emptyset$. 
\item[$\bullet$]
The $x_j$ here are $state\ variables$ of state $s'$.

@@

A composite pNet consists of a set of sub-pNets exposing
a set of actions, each of them triggering internal actions in each of
the sub-pNets. The synchronization between global actions and
internal actions is given by  \emph{synchronization vectors}: a
synchronization vector synchronizes one or several internal actions, and
exposes a single resulting global action.
% Actions involved at the pNet level (in the synchronization vectors) do
% not need to distinguish between input and output
% variables.
%Action terms for pNets are defined as follows:
%\[\begin{array}[l]{rcl@{\quad}l}
%  \alpha\in \AlgAS &::=&a(Expr_1,\ldots,Expr_n)
%\end{array}
%\]

\begin{definition}[pNets]\label{def-pnets}
A pNet is a hierarchical structure which leaves are pLTSs and holes:\\
$\pNet\triangleq pLTS~|~\mylangle \pNet_i^{i\in I}, S_j^{j\in J}, \symb{SV}_k^{k\in K}\myrangle$
where
\begin{itemize}
\item[$\bullet$] $I \in \I$ is the set over which sub-pNets are indexed.
\item[$\bullet$] $\pNet_i^{i\in I}$ is the family of sub-pNets.
%  $\pNet_i^{i\in I}$ is a family of sub-pNets where $I\in\I_\P$ is the set over which sub-pNets are indexed.

@@

\section{Running Example}
\label{section:examples}

@@

 \begin{figure}[t]
   \centerline{\includegraphics[width=10cm]{XFIG/LOTOSFormula.eps}}
   \caption{pNet encodings for the LOTOS expression}  \label{schema:lotos-pnet}
 \end{figure}

We use an example inspired from LOTOS, to illustrate
the main features of pNets and our semantic construction algorithm.
%% \QIN{
%% Meanwhile, 
%% the formula using operators from CCS can also be encoded into a pNet and 
%% conducted with the same things.
%% }
%% \subsection{LOTOS Formula Example}
We choose here the expression \texttt{P>>(acc(v);Q)}, that includes the
Enable and Prefix 
operators of the LOTOS language. For the readers not familiar with
Lotos, the pNet encodings here should be sufficient to understand the
Enable operator semantics.
In Fig.~\ref{schema:lotos-pnet}, we show the encoding of the formula using pNets. 
This pNet is composed of: a toplevel node PN1 encoding the Enable
operator, a pLTS C1 acting as the {\sl controller} of the node; a hole
representing the 
process variable P; another node PN2 representing the Prefix
operator, with its controller C2, and its hole Q.

In order to show all features of the pNet model, we choose to use two
styles of encodings in the controllers.
% The most common style used in process calculi is ``state oriented'',
Controller C2 is ``state oriented'': it has two states to
indicate the position of the control point in the
Prefix expression $(acc(v);Q)$. The second one is
``data oriented'': controller C1 has a single state, with a state
variable $s_0$ encoding the change of control point from the left to
the right subnet.
%% The transitions in the pLTS have the more general syntax:
%% ``<action-expr> [guard] {sequence of assignments}''. Guards and
%% assignments can only use the values of variables of their source state
%% and values from input variables in the action expression; and a
%% transition can only assign variables of its target state.

The root pNet node PN1 has 3 sub-pNets: its controller C1,
the hole P and the sub-pNet PN2.
Three synchronization vectors define the possible
synchronizations between the actions of the subnets.
The first vector ``$< l, a_1 , \_ > -> a_1$'' transmits an action
$a_1$ of P to the upper level. Here the ``l'' action is related
to the corresponding transition of the controller C1, ``$\_$'' means
that the corresponding subnet is not involved in this synchronization,
and there is a guard meaning this synchronisation can only occur if
$a_1$ is not a $\delta$.
The second synchronisation vector composes an
action $\delta(x)$ from $P$ with an $acc(x)$ from $Q$, transmitting a
value $x$ before $Q$ takes the control; 
The vector result $\underline{\delta(x)}$ is a ``synchronized''
action, that cannot be further
synchronized at upper levels. Synchronized actions is a useful
generalization of the notion of internal actions, that will be
convenient for observing internal events during model-checking.
% The last vector acts is to the first one.  
% Similarly the node PN2 has two synchronization vectors to encode
% the behavior of the Prefix operator.

It should be easy to see that this pNet system encodes properly the
behavioral semantics of the corresponding algebra expression. We remark
that it is built in a systematic structural way, one pNet node
encoding each operator.
%% It is parameterized both at the level of
%% actions variables (occurring within the controllers and the
%% synchronization vectors), and at the level of process parameters
%% (holes).

%% \TODO{I'm thinking about moving this second example to section 5...}
%% \QIN{[This CCS example is removed.]}

%\subsection{CCS Parallel Example}
%
%\begin{figure}[h]
%  \centerline{\includegraphics[width=10cm]{XFIG/CCS1.eps}}
%  \caption{pNet encodings for the CCS expression}  \label{schema:ccs-pnet}
%\end{figure}
%
%\QIN{
%This CCS expression composing two operators, parallel and prefix. In Fig.~\ref{schema:ccs-pnet}, 
%it is easy to recognize that the pNets $PN2$ and $PN3$ are similar to 
%the sub-pNet in the LOTOS example in their structure and synchronization vectors. 
%That is because they are the operators conduct a similar behavior from different algebras.
%}
%
%\QIN{
%To present the parallel property, we have defined a set of synchronization vectors composing 
%two subnets. We use the notation from the CCS to present the emit and receive action. 
%The variables $v_1$, $v_2$ are the data transmitting between two components, 
%showing its value-passing behavior.
%}

@@

\begin{figure}[h]
  \centerline{\includegraphics[width=5cm]{XFIG/LOTOSresult}}

>>>>

\begin{abstract}
  \TODO{not good for a FM, (to be rewriten by Eric)}

@@

\paragraph{Contribution}
In the article we show how:
\begin{itemize}
\item We define the open automaton generation algorithm, and we
  implemented a full working prototype, within the 
  VerCors platform. In the process, we improved the
  semantics rules from~\cite{henrio:Forte2016}, and add features in
  the algorithm to deal the full 
  model, including management of variables and assignments.
%  \item We extended the meta-model of action expressions, and the
%    associated API, in the VerCors platform, to be able to manage
%    parameterisation over the action algebra operators.
\item We implement the interaction between our algorithm and the Z3
    SMT solver, for checking satisfiability of the transitions
    generated by the algorithm.
  \item \ERIC{We show the interest of this approach on an
    industrial-inspired use-case, namely one architectural pattern
    extracted from the BIP specification of a nano satellite on-bord
    software.} 
\end{itemize}

@@

\TODO{To be reduced, moving more material to the appendix}

@@

pNets can encode naturally the notion of input actions as found e.g. in value-passing CCS
\cite{Milner89} or of usual point-to-point message passing calculi, but it also allows
for more general mechanisms, like gate negotiation in Lotos, or broadcast
communications.

@@

\begin{definition}[pLTS]
\label{pLTS}
A pLTS is a tuple
$pLTS\triangleq\mylangle S,s_0, \to\myrangle$ where:
\begin{itemize}
\item[$\bullet$]
$S$ is a set of states.
\item[$\bullet$]
$s_0 \in S$ is the initial state.
\item[$\bullet$] $\to \subseteq S \times L \times S$ is the transition relation and 
$L$ is the set of labels of the form
$\langle \alpha,~e_b,~(x_j\!:= {e}_j)^{j\in J}\rangle$,
where $\alpha \in\AlgA$ is a parameterized action, $e_b \in
\AlgB$ is a guard, and the variables $x_j\in P$
are assigned the expressions $e_j\in \AlgE$.
If 
$s \xrightarrow{\langle \alpha,~e_b,~(x_j\!:= {e}_j)^{j\in
		J}\rangle} s'\in \to $ then 
		$\vars(\alpha)\!\subseteq\! \vars(s)$, 
		$\vars(e_b)\!\subseteq\! \vars(s')$, and
		$\forall j\!\in\! J .\,\vars(e_j)\!\subseteq\! \vars(s)\land 
x_j\!\in\!\vars(s')$.
\item[$\bullet$]
The $x_j$ here are $state\ variables$ of state $s'$.

@@

A composite pNet consists of a set of sub-pNets exposing
a set of actions, each of them triggering internal actions in each of
the sub-pNets. The synchronization between global actions and
internal actions is given by  \emph{synchronization vectors}: a
synchronization vector synchronizes one or several internal actions, and
exposes a single resulting global action.

\begin{definition}[pNets]\label{def-pnets}
A pNet is a hierarchical structure which leaves are pLTSs and holes:\\
$\pNet\triangleq pLTS~|~\mylangle \pNet_i^{i\in I}, S_j^{j\in J}, \symb{SV}_k^{k\in K}\myrangle$
where
\begin{itemize}
\item[$\bullet$] $I \in \I$ is the set over which sub-pNets are indexed.
\item[$\bullet$] $\pNet_i^{i\in I}$ is the family of sub-pNets.

@@

\section{BIP architectures, and their encodings into pNets}
\label{section:BIParchitectures}

\section{Running example}
\label{section:FailureTimerMax}

@@

 \begin{figure}[t]
   \centerline{\includegraphics[width=12cm]{XFIG/BIPspec-ArchFailureTimerMax}}
   \caption{FailureTimer Architecture in BIP graphical syntax}  \label{schema:BIParchitecture}
 \end{figure}

\begin{figure}[t]
   \centerline{\includegraphics[width=12cm]{XFIG/BIPpnet-ArchFailureTimerMax}}
   \caption{FailureTimer pNet encoding}  \label{schema:BIParchitecture}
 \end{figure}

@@

\begin{figure}[t]
   \centerline{\includegraphics[width=12cm]{XFIG/BIPresult-ArchFailureTimerMax}}
   \caption{Open Automaton\TODO{syntax could be better...}}  \label{schema:resultOA}
 \end{figure}

\begin{figure}[h]
%  \centerline{\includegraphics[width=5cm]{XFIG/LOTOSresult}}
